### 2018-01-04 manuscript.rmd
finished pass at edits in main text

New data were analysed in conjunction with 13 publicly available datasets, with a total number of 43 species. Sequence assembly, annotation, homology evaluation, gene tree construction, parsing of genes trees to isolate orthologous sequences, and supermatrix construction were conducted with Agalma v. 1.00 [@Dunn:2013kw; @Guang202416].  This workflow integrates a variety of existing tools [@grabherr2011full, @altschul1990basic, @doi:10.1093/nar/30.7.1575, @stamatakis2006raxml, @sukumaran2010dendropy, @langmead2012fast, @li2011rsem, @talavera2007improvement, @li2009sequence, @katoh2013mafft] and new methods. Maximum likelihood analyses of the supermatrix were conducted with [XX add raxml version and citation]. Bayesian Inference (BI) analyses of the supermatrix were conducted using Phylobayes v. 1.7a-mpi [@lartillot2009phylobayes]. Sequence alignments, sampled and consensus trees, and voucher information are available in the git repository.

@@

![Phylogram of siphonophore relationships. Node labels indicate bootstrap support percent, unnumbered nodes have 100% support. The image was rendered with ggtree [@Yu:2016fo].](figures/Figure3.pdf)

Using the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic], we evaluated the following three alternative phylogenetic hypotheses against the most likely tree topology (Fig. 3C): (i) monophyletic Physonectae, (ii) monoecious siphonophores (not including *Rudjakovia* sp.) are monophyletic, and (iii) monoecious siphonophores (including *Rudjakovia* sp.) are monophyletic. We performed tests (ii) and (iii) because it is not known if *Rudjakovia* sp. is monoecious or dioecious. In all three tests the alternative hypothesis was rejected (p-value <0.01, confidence interval: <0.001 - 0.03, Fig. S2).

@@

In all siphonophores, each gonophore (sexual medusa that produces gametes) is either male or female. Within each siphonophore species, colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). The current better-resolved tree (Fig. 5A), indicates that monoecy in siphonophores evolved twice from a dioecious ancestor, in the branch leading to Calycophorae and in the branch leading to Clade A. There is a small probability for an alternative scenario featuring a single gain of monoecy before the split of Codonophora, with a subsequent shift to dioecy in the *Marrus-Erenna* clade. Previous analyses suggested that the common ancestor of siphonophores was dioecious, and there is a single gain of monoecy within the Codonophora [@Dunn:2005dy].

Calycophorae and Clade A each have very different arrangements of male and female gonophores, consistent with the multiple gains of monoecy supported here. In calycophorans, each cormidium (a single reiterated sequence of zooids along the stem, see Fig. 2) bears either male or female zooids. In this form of monoecy, the male and female cormidia can either be in an alternating pattern, or there can be several male or female cormidia in a row. In either case, male and female zooids are found at the same corresponding locations within the cormidia. In Clade A, however, male and female zooids are found within the same cormidium. In these species, the male and female zooids are placed at very different but well defined locations within the cormidium.

@@

One of the most striking aspects of siphonophore biology is their diversity of unique zooid types [@Beklemishev1969;@cartwright2010character]. For example, members of *Forskalia* have 5 basic zooid types (nectophore, gastrozooid, palpon, bract, and gonophore), and in some species, a total of 9 when considering zooid subtypes (4 types of bract, male & female gonophores)[@pugh2003revision]. Here we reconstruct the evolutionary origins of several zooid types on the present transcriptome tree (Fig. 4).

Nectophores are non-reproductive propulsive medusae that codonophorans use for coordinated colony-level swimming (Fig. 2). Nectophores are also present within the gonodendra (reproductive structures) of the cystonects, and are thought to propel the gonodendra when they detach from the colony [@totton1965synopsis;@Totton1960]. It is not clear whether the nectophores found within the siphosome of the cystonects are homologous to the nectophores borne on the nectosome of codonophorans. The nectosome is the region of the codonophoran colony that develops from the nectosomal growth zone (Fig. 2B). The nectosomal growth zone buds nectophores (and in the case of *Apolemia*, also palpons). The present tree, as well as the analyses of @Dunn:2005dy, are consistent with a single origin of the nectosome (Fig. S5).

@@

The nectosome can be located in a dorsal or a ventral position, following the orientation framework of @haddock_dunn_pugh_2005. Our ancestral reconstructions for this character (Fig. S8) show that a ventrally-oriented nectosome was the ancestral form in Codonophora, and that a dorsal nectosome has evolved twice independently, in the branch leading to Agalmatidae and in *Bargmannia*.

@@

The most complex nematocyst batteries of Cnidaria can arguably be found among the siphonophores, hanging in regularly spaced tentacle side branches called tentilla (Fig. 2A). Most hydrozoans bear simple tentacles (tentacles with no side branches). It is still an open question whether the common ancestor of Siphonophora had tentilla. The only siphonophores species regarded as lacking tentilla are *Physalia physalis*, *Apolemia* spp., and *Bathyphysa conifera*. Since *B. conifera* is the only member of the *Rhizophysidae* (and of the *Bathyphysa* genus) lacking tentilla, we can safely assume this is a case of secondary loss. When we reconstruct the evolution of this character on the current phylogeny, 70% of simulations support an common ancestor bearing tentilla, with two independent losses leading to *Physalia* and *Apolemia* (Fig. S3). However, this leaves a 30% support for a simple-tentacled common ancestor followed by 2 independent gains of tentilla in the branches leading to *Rhizophysidae* and non-apolemiid codonophorans.

@@

Siphonophores are abundant predators in the pelagic realm, ranging from the surface (*Physalia physalis*) to bathypelagic depths (Fig. 4, S9) [@mapstone2014global; @mackie1987siphonophore]. However, the depth distribution of siphonophore populations is not always static. Many species are strong vertical migrators, and some species such as *Chuniphyes multidentata* have exhibited synchronous diel migration patterns [@pugh1984diel]. Some siphonophore species (such as *Bargmannia lata*) have been observed at abyssopelagic depths near 4000m. Using the present phylogeny, we reconstructed the median depth changes along the phylogeny under a Brownian Motion model (Fig. S9), which had the strongest AICc support (compared to non-phylogenetic distributions, and to Ohrnstein-Uhlenbeck). This model indicates a mesopelagic most recent common ancestor, with several independent transition events to epipelagic (light blue) and bathypelagic (dark blue) waters. This reconstruction only included species recorded using an ROV, thus it excludes many other independent colonizations of the epipelagic habitat. The ROV observations are reliable below 200m, and no quantative measurements were made on SCUBA dives. Species such as *Hippopodius hippopus*, *Athorybia rosacea*, *Diphyes dispar*, and *Chelophyes appendiculata* are often encountered blue water diving less than 20m from the surface (Fig. 4). We also reconstructed the median depth changes along the phylogeny using median depths of 20m for all species collected by SCUBA diving (Fig. S10), and still find support for a mesopelagic ancestor. There was only a single transition to benthic lifestyle on the branch of *Rhodaliidae*, and a single transition to a pleustonic lifestyle on the branch of *Physalia physalis*.

>>>>

New data were analysed in conjunction with 13 publicly available datasets, with a total number of 43 species. Sequence assembly, annotation, homology evaluation, gene tree construction, parsing of genes trees to isolate orthologous sequences, and supermatrix construction were conducted with Agalma v. 1.00 [@Dunn:2013kw; @Guang202416].  This workflow integrates a variety of existing tools [@grabherr2011full, @altschul1990basic, @doi:10.1093/nar/30.7.1575, @stamatakis2006raxml, @sukumaran2010dendropy, @langmead2012fast, @li2011rsem, @talavera2007improvement, @li2009sequence, @katoh2013mafft] and new methods. Maximum likelihood analyses of the supermatrix were conducted with [XX add raxml version and citation]. Bayesian Inference (BI) analyses of the supermatrix were conducted using Phylobayes v. 1.7a-mpi [@lartillot2009phylobayes]. Sequence alignments, sampled and consensus trees, and voucher information are available in the git repository. Tree figures were rendered with ggtree [@Yu:2016fo].

@@

![(A) Maximum likelihood (ML) phylogram with bipartation frequencies from the ML bootstraps and the Bayesian posterior distribution of trees. Unlabeled nodes have support >0.99 for both bootstraps and posteriors. (B) The topologies found in the psoterior distribution of trees that conflict with the ML tree. (C) The topologies evaluated by the SOWH tests. For more details on the SOWH topologies refer to Fig. S2.](figures/Figure3.pdf)

Using the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic], we evaluated the following three alternative phylogenetic hypotheses against the most likely tree topology (Fig. 3C): (i) monophyletic Physonectae, (ii) monophyletic monoecious siphonophores (both with and without *Rudjakovia* sp.). In all three tests the alternative hypothesis was rejected (p-value <0.01, confidence interval: <0.001 - 0.03, Fig. S2).

@@

In all siphonophores, each gonophore (sexual medusa that produces gametes) is either male or female. Within each siphonophore species, colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analyses suggested that the common ancestor of siphonophores was dioecious, and was consistent with a single gain of monoecy within the Codonophora and no secondary losses [@Dunn:2005dy]. The better-resolved tree in the current analyses indicates that the evolution of monoecy is more complicated than this. The two clades of monophyloetic siphonophores, Calycophorae and Clade A (Fig. 3A), do not form a monophyletic group. This is because Clade B, which contains dioecious species, is also desceneded from their most recent common ancestor. SOWH tests strongly reject the placement of the monecious clades Calycophorae and Clade A as a group that excludes Clade B (Figs. 3C and S2). The positions of the only two taxa from Clade B in the previous analysies [@Dunn:2005dy], *Erenna* and *Stephalia*, were unresolved in that previous study. This difference in conclusions regarding trait evolution therefore does not reflect a contradication between alternative strongly supported results, but the resolution of earlier polytomies in a way that indicates there has been homoplasy in the evolution of monoecy.

The distribution of monoecy is consistent with two scenarios. In the first, there is a single shift from dioecy to monoecy along the branch that give rise to the most recont common ancestor of Calycophorae and Euphysonectae, followed by a shift back to dioecy along the branch that gave rise to Clade B. In the second, monoecy arose twice -- once along the branch that gave rise to Clade A and again along the branch that gave rise to Calycophorae.

Ancestral character state reconstructions favor the hypothesis that monoecy arose twice (Fig. 5A), once in Calycophorae and once in Clade A. This is consistent with differences in the arrangements of male and female gonophores in the two clades. In calycophorans, each cormidium (a single reiterated sequence of zooids along the stem, see Fig. 2) bears either male or female zooids. In this form of monoecy, the male and female cormidia can either be in an alternating pattern, or there can be several male or female cormidia in a row. In either case, male and female zooids are found at the same corresponding locations within the cormidia. In Clade A, however, male and female zooids are found within the same cormidium. In these species, the male and female zooids are placed at very different but well defined locations within the cormidium.

@@

One of the most striking aspects of siphonophore biology is their diversity of unique zooid types [@Beklemishev1969; @cartwright2010character]. For example, members of *Forskalia* have 5 basic zooid types (nectophore, gastrozooid, palpon, bract, and gonophore), and in some species, a total of 9 when considering zooid subtypes (4 types of bract, male & female gonophores)[@pugh2003revision]. Here we reconstruct the evolutionary origins of several zooid types on the present transcriptome tree (Fig. 4).

Nectophores (Fig. 2) are non-reproductive propulsive medusae. In Codonophora, the nectophores are localized to a region known as the nectosome (Fig. 2B), which has its own growth zone, and are used for coordinated colony-level swimming. Nectophores are also present within the gonodendra (reproductive structures) of cystonects, and are thought to propel the gonodendra when they detach from the colony [@totton1965synopsis;@Totton1960]. It is not clear whether the nectophores found within the siphosome of the cystonects are homologous to the nectophores borne on the nectosome of codonophorans. The present analyses, as well as the analyses of @Dunn:2005dy, are consistent with a single origin of the nectosome (Fig. S5).

@@

Within the nectosome, the nectophores can be attached along the dorsal or ventral side of the stem, following the orientation framework of @haddock_dunn_pugh_2005. Our ancestral reconstructions for this character (Fig. S8) suggest that ventral attachment of nectophores was the ancestral state in Codonophora, and that dorsal attachment hasindependently evolved twice -- once along the Agalmatidae stem and once along the Pyrostephidae stem.

@@

The most complex nematocyst batteries of Cnidaria are found among the siphonophores, hanging in regularly spaced tentacle side branches called tentilla (Fig. 2A). Most hydrozoans bear simple tentacles (tentacles with no side branches). It is still an open question whether the common ancestor of Siphonophora had tentilla. The only siphonophores species regarded as lacking tentilla are *Physalia physalis*, *Apolemia* spp., and *Bathyphysa conifera*. Since *B. conifera* is the only member of the *Rhizophysidae* (and of the *Bathyphysa* genus) lacking tentilla, we can safely assume this is a case of secondary loss. When we reconstruct the evolution of this character on the current phylogeny, 70% of simulations support an common ancestor bearing tentilla, with two independent losses leading to *Physalia* and *Apolemia* (Fig. S3). However, this leaves a 30% support for a simple-tentacled common ancestor followed by 2 independent gains of tentilla in the branches leading to *Rhizophysidae* and non-apolemiid codonophorans.

@@

Siphonophores are abundant predators in the pelagic realm, ranging from the surface (*Physalia physalis*) to bathypelagic depths (Figs. 4 and S9) [@mapstone2014global; @mackie1987siphonophore]. However, the depth distribution of siphonophore populations is not always static. Many species are strong vertical migrators, and some species such as *Chuniphyes multidentata* have exhibited synchronous diel migration patterns [@pugh1984diel]. Some siphonophore species (such as *Bargmannia lata*) have been observed at abyssopelagic depths near 4000m. Using the present phylogeny, we reconstructed the median depth changes along the phylogeny under a Brownian Motion model (Fig. S9), which had the strongest AICc support (compared to non-phylogenetic distributions, and to Ohrnstein-Uhlenbeck). This model indicates a mesopelagic most recent common ancestor, with several independent transition events to epipelagic (light blue) and bathypelagic (dark blue) waters. This reconstruction only included species recorded using an ROV, thus it excludes many other independent colonizations of the epipelagic habitat. The ROV observations are reliable below 200m, and no quantative measurements were made on SCUBA dives. Species such as *Hippopodius hippopus*, *Athorybia rosacea*, *Diphyes dispar*, and *Chelophyes appendiculata* are often encountered blue water diving less than 20m from the surface (Fig. 4). We also reconstructed the median depth changes along the phylogeny using median depths of 20m for all species collected by SCUBA diving (Fig. S10), and still find support for a mesopelagic ancestor. There was only a single transition to benthic lifestyle on the branch of *Rhodaliidae*, and a single transition to a pleustonic lifestyle on the branch of *Physalia physalis*.
